---
title: "Environmental Toxic Burden, Socioeconomic Deprivation, and Mortality Inequality"
subtitle: "A Causal Analysis Across Three Levels of Inference | Pooled NHANES 2007–2012"
author: "Kaung Khant Tin"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

---

# Abstract {.unnumbered}

This analysis uses pooled data from three cycles of the US National Health and Nutrition Examination Survey (NHANES 2007–2012; N = 10,410 White and Black Americans; 1,780 deaths) to investigate whether blood cadmium — a biomarker of cumulative environmental toxic burden — partially explains the well-established link between socioeconomic deprivation and premature mortality. The analysis is structured around Judea Pearl's three levels of causal inference: association, intervention, and counterfactual. After adjusting for age, sex, race, socioeconomic status, and smoking, each e-fold increase in blood cadmium was associated with 36% higher all-cause mortality (HR 1.364, 95% CI 1.247–1.492). Importantly, the crude racial mortality gap disappeared entirely once these structural pathways were accounted for (HR 0.938, p = 0.282), suggesting that racial health inequalities in this context are structurally produced rather than biologically inherent. Using g-computation, the analysis estimates that equalising cadmium exposure between the most and least deprived groups would improve 10-year survival by 1.62 percentage points (95% CI 1.14–2.14), corresponding to approximately 12.85% of excess mortality among people below the poverty line. These findings provide a direct methodological and strong foundation for the proposed Manchester PhD project on environmental causes of cancer care inequalities, where the same causal inference approach can be applied to UK cancer registry data linked with spatially resolved pollutant measurements.

---

# Background and Motivation

## Context

Socioeconomic deprivation is one of the most consistently documented predictors of premature mortality, yet the specific causal pathways through which it leads to earlier death remain only partially understood. One pathway that has received comparatively little attention is **environmental toxic burden** — the cumulative exposure to industrial, occupational, and dietary pollutants that tends to be systematically higher in more deprived communities. This pattern arises because residential segregation, land-use planning decisions, and occupational structures mean that poorer populations are more likely to live near pollution sources, work in exposure-prone industries, and consume food grown in contaminated soils.

**Cadmium** is a heavy metal classified as a Group 1 carcinogen by the International Agency for Research on Cancer (IARC), with well-established toxic effects on the kidneys and multiple organ systems. Blood cadmium serves as a useful biomarker of cumulative environmental toxic burden because it captures exposure from multiple sources simultaneously: cigarette smoke, industrial emissions, contaminated food, and occupational contact. A key biological property makes cadmium particularly informative — once absorbed, it accumulates in the renal tubules over decades and the body has no effective mechanism for eliminating it. This means that even a single blood measurement provides a reasonable approximation of an individual's lifetime environmental burden.

This analysis examines whether cadmium partially explains (mediates) the relationship between socioeconomic deprivation and all-cause mortality. It is structured around Judea Pearl's three-level framework for causal inference, sometimes called the "ladder of causation": **association** (Level 1 — what patterns do we observe?), **intervention** (Level 2 — what happens when we adjust for confounders?), and **counterfactual** (Level 3 — what would have happened under a different exposure scenario?). This framework was chosen deliberately because the proposed PhD project at Manchester's Division of Cancer Sciences centres on causal inference methods for understanding environmental determinants of cancer care inequalities — the same methodological toolkit applied here to a different data context.

## Research Questions

**Primary question:** Is higher blood cadmium independently associated with all-cause mortality after adjusting for age, sex, race, socioeconomic status (SES), and smoking?

**Secondary questions:**

1. Does SES confound the cadmium-mortality relationship, and if so, in which direction?
2. Does the cadmium-mortality association vary by race after full adjustment?
3. Do Black Americans carry higher cadmium than White Americans even within the same smoking categories — suggesting non-smoking environmental sources?
4. Does adjusting for smoking attenuate the cadmium effect, and if so, by how much?
5. Does race modify the effect of cadmium on mortality, or is the effect consistent across racial groups?
6. What proportion of excess mortality among people below the poverty line can be attributed to their higher cadmium exposure?

## Dataset

This analysis brings together four separate data sources from the US Centers for Disease Control and Prevention (CDC), linked at the individual level using each participant's unique sequence number (SEQN):

| Source | Content |
|---|---|
| NHANES Demographics (DEMO) | Age, sex, race/ethnicity, poverty-income ratio, survey weights |
| NHANES Laboratory — PBCD | Blood cadmium (μg/L) |
| NHANES Laboratory — COTNAL | Serum cotinine (ng/mL) — an objective biomarker of tobacco smoke exposure |
| NHANES Smoking Questionnaire — SMQ | Self-reported smoking behaviour |
| CDC Mortality Linkage Files | All-cause and cancer mortality status, follow-up time via the National Death Index |

**Why pool three survey cycles?** A single NHANES cycle provides too few events for stable statistical modelling — for example, only approximately 23 cancer deaths among Black Americans in one cycle, which falls well below the minimum of 10 events per variable needed for reliable regression estimates. Pooling three consecutive cycles (2007–08, 2009–10, and 2011–12) produces a combined sample of N = 10,410 White and Black Americans with 1,780 all-cause deaths, providing adequate statistical power. Following CDC guidelines for pooling across cycles, the 2-year survey weights are divided by 3.

---

# Data Download and Preparation

## Packages
```{r packages}
required_packages <- c("nhanesA", "tidyverse", "survival",
                       "gtsummary", "survminer", "readr",
                       "broom", "patchwork", "knitr")

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  }
}
```

## Download Mortality Files

Mortality linkage files are fixed-width text files from the CDC FTP server. We use `read_fwf` with exact column positions from the CDC codebook, downloaded into a temporary file — nothing written to disk.
```{r mortality}
read_nhanes_mort <- function(url) {
  mort_file <- tempfile()
  download.file(url, mort_file, mode = "wb", quiet = TRUE)
  read_fwf(
    mort_file,
    fwf_cols(
      SEQN         = c(1,  6),
      eligstat     = c(15, 15),
      mortstat     = c(16, 16),
      ucod_leading = c(17, 19),
      diabetes     = c(20, 20),
      hyperten     = c(21, 21),
      permth_int   = c(43, 45),
      permth_exm   = c(46, 48)
    ),
    show_col_types = FALSE
  )
}

base_url <- paste0("https://ftp.cdc.gov/pub/Health_Statistics/NCHS/",
                   "datalinkage/linked_mortality/")

mort_2007 <- read_nhanes_mort(paste0(base_url,
  "NHANES_2007_2008_MORT_2019_PUBLIC.dat"))
mort_2009 <- read_nhanes_mort(paste0(base_url,
  "NHANES_2009_2010_MORT_2019_PUBLIC.dat"))
mort_2011 <- read_nhanes_mort(paste0(base_url,
  "NHANES_2011_2012_MORT_2019_PUBLIC.dat"))

cat("Mortality records downloaded:\n")
cat("  2007-08:", nrow(mort_2007), "\n")
cat("  2009-10:", nrow(mort_2009), "\n")
cat("  2011-12:", nrow(mort_2011), "\n")
```

## Download Survey Files

NHANES laboratory and questionnaire files downloaded via `nhanesA`, which handles all XPT parsing directly from CDC servers.
```{r survey-files}
# 2007-08
demo_2007    <- nhanes("DEMO_E")
lab_cdm_2007 <- nhanes("PBCD_E")
lab_cot_2007 <- nhanes("COTNAL_E")
smq_2007     <- nhanes("SMQ_E")

# 2009-10
demo_2009    <- nhanes("DEMO_F")
lab_cdm_2009 <- nhanes("PBCD_F")
lab_cot_2009 <- nhanes("COTNAL_F")
smq_2009     <- nhanes("SMQ_F")

# 2011-12
demo_2011    <- nhanes("DEMO_G")
lab_cdm_2011 <- nhanes("PBCD_G")
lab_cot_2011 <- nhanes("COTNAL_G")
smq_2011     <- nhanes("SMQ_G")

cat("Survey files downloaded for all three cycles.\n")
```

## Harmonise and Pool

Three non-trivial harmonisation challenges arose when pooling cycles:

1. **Race variable naming.** `RIDRETH1` (2007-08, 2009-10) vs `RIDRETH3` (2011-12). Detected and handled within the cleaning function.
2. **Race coding.** `nhanesA` returns labelled text values (e.g. "Non-Hispanic White") rather than numeric codes — the `case_when` maps these to consistent labels.
3. **Smoking.** Self-reported smoking (`SMQ040`) joined from a separate questionnaire file via left join — participants without a smoking record are classified as Never.
```{r harmonise}
clean_cycle <- function(demo, mort, cdm, cot, smq, cycle_year) {

  # Race variable changed name in 2011-12
  if ("RIDRETH3" %in% names(demo)) {
    demo <- demo %>% mutate(RACE_VAR = as.character(RIDRETH3))
  } else {
    demo <- demo %>% mutate(RACE_VAR = as.character(RIDRETH1))
  }

  demo %>%
    inner_join(mort, by = "SEQN") %>%
    inner_join(cdm %>% select(SEQN, LBXBCD), by = "SEQN") %>%
    inner_join(cot %>% select(SEQN, LBXCOT), by = "SEQN") %>%
    left_join(smq  %>% select(SEQN, SMQ040), by = "SEQN") %>%
    filter(
      eligstat == 1,
      RIDAGEYR >= 20,
      mortstat %in% c("0", "1")
    ) %>%
    mutate(
      cycle          = cycle_year,
      surv_months    = as.numeric(permth_exm),
      allcause_death = as.numeric(mortstat),

      cancer_death = case_when(
        mortstat == "1" & ucod_leading == "002" ~ 1,
        mortstat == "0"                         ~ 0,
        mortstat == "1" & ucod_leading != "002" ~ 0,
        TRUE ~ NA_real_
      ),

      race = case_when(
        RACE_VAR == "Non-Hispanic White"                    ~ "White",
        RACE_VAR == "Non-Hispanic Black"                    ~ "Black",
        RACE_VAR %in% c("Mexican American", "Other Hispanic",
                        "Hispanic")                         ~ "Hispanic",
        RACE_VAR == "Non-Hispanic Asian"                    ~ "Asian",
        TRUE ~ NA_character_
      ),
      race = factor(race, levels = c("White", "Black",
                                     "Hispanic", "Asian")),

      sex           = factor(RIAGENDR,
                             levels = c("Male", "Female")),
      age           = RIDAGEYR,
      poverty_ratio = INDFMPIR,

      ses_cat = case_when(
        INDFMPIR <  1.0 ~ "Below poverty",
        INDFMPIR <  3.0 ~ "Low-middle",
        INDFMPIR >= 3.0 ~ "Middle-high",
        TRUE ~ NA_character_
      ),
      ses_cat = factor(ses_cat,
                       levels = c("Middle-high", "Low-middle",
                                  "Below poverty")),

      education = case_when(
        DMDEDUC2 %in% c(
          "Less than 9th grade",
          "9-11th grade (Includes 12th grade with no diploma)"
        ) ~ "Less than high school",
        DMDEDUC2 == "High school graduate/GED or equivalent"
          ~ "High school",
        DMDEDUC2 %in% c("Some college or AA degree",
                        "College graduate or above")
          ~ "College or above",
        TRUE ~ NA_character_
      ),
      education = factor(education,
                         levels = c("College or above", "High school",
                                    "Less than high school")),

      cadmium_raw  = LBXBCD,
      cadmium_log  = log(LBXBCD),
      cotinine_raw = LBXCOT,
      cotinine_log = log(LBXCOT + 0.011),

      smoking = case_when(
        SMQ040 == "Every day"  ~ "Current",
        SMQ040 == "Some days"  ~ "Current",
        SMQ040 == "Not at all" ~ "Former/Never",
        is.na(SMQ040)          ~ "Never",
        TRUE ~ NA_character_
      ),
      smoking = factor(smoking,
                       levels = c("Never", "Former/Never", "Current"))
    ) %>%
    filter(
      !is.na(surv_months), !is.na(cancer_death),
      !is.na(cadmium_raw), !is.na(cotinine_raw),
      !is.na(race), !is.na(age), !is.na(sex),
      surv_months > 0
    )
}

cycle_2007 <- clean_cycle(demo_2007, mort_2007,
                          lab_cdm_2007, lab_cot_2007,
                          smq_2007, "2007-2008")
cycle_2009 <- clean_cycle(demo_2009, mort_2009,
                          lab_cdm_2009, lab_cot_2009,
                          smq_2009, "2009-2010")
cycle_2011 <- clean_cycle(demo_2011, mort_2011,
                          lab_cdm_2011, lab_cot_2011,
                          smq_2011, "2011-2012")

cat("Per-cycle merge complete:\n")
cat("  2007-08:", nrow(cycle_2007), "\n")
cat("  2009-10:", nrow(cycle_2009), "\n")
cat("  2011-12:", nrow(cycle_2011), "\n")
```

## Pool and Finalise
```{r pool}
pooled_data <- bind_rows(cycle_2007, cycle_2009, cycle_2011) %>%
  mutate(
    cadmium_q = ntile(cadmium_raw, 4),
    cadmium_q = factor(cadmium_q, levels = 1:4,
                       labels = c("Q1 (lowest)", "Q2",
                                  "Q3", "Q4 (highest)")),
    cadmium_hilow = factor(
      ifelse(cadmium_raw >= median(cadmium_raw, na.rm = TRUE),
             "High", "Low"),
      levels = c("Low", "High")
    ),
    wtmec_pooled = WTMEC2YR / 3
  )

# White + Black analytical subset
wb_data <- pooled_data %>%
  filter(race %in% c("White", "Black")) %>%
  mutate(race = droplevels(race))

# Survival objects
surv_allcause_wb <- Surv(wb_data$surv_months, wb_data$allcause_death)
surv_cancer_wb   <- Surv(wb_data$surv_months, wb_data$cancer_death)
surv_cancer_all  <- Surv(pooled_data$surv_months, pooled_data$cancer_death)

cat("=====================================================\n")
cat("POOLED NHANES DATASET READY\n")
cat("=====================================================\n")
cat("Total participants (all races):", nrow(pooled_data), "\n")
cat("White + Black subset:          ", nrow(wb_data),     "\n")
cat("All-cause deaths (W+B):        ",
    sum(wb_data$allcause_death), "\n")
cat("Cancer deaths (W+B):           ",
    sum(wb_data$cancer_death),   "\n")
cat("Event rate:                    ",
    round(mean(wb_data$allcause_death)*100, 1), "%\n")
cat("Follow-up range (months):      ",
    min(wb_data$surv_months), "–", max(wb_data$surv_months), "\n")
```

---

# Conceptual Framework

## Directed Acyclic Graph (DAG)

Before fitting any statistical models, it is essential to make our causal assumptions explicit. A Directed Acyclic Graph (DAG) is a visual tool that maps out the assumed relationships between variables — it tells us which variables we need to adjust for, which we should leave alone, and what confounding or mediation structures we expect. Getting this step right is arguably more important than the statistical modelling itself, because no amount of sophisticated analysis can compensate for a poorly specified causal structure.

The DAG below represents the assumed causal relationships in this analysis:

```
Residential segregation ──→ Race ──→ Neighbourhood deprivation
                                             │
                                             ↓
                             SES ──────→ Cadmium ──────────→ Mortality
                              │                                   ↑
                              └───────────────────────────────────┘
                                      (direct SES pathway)

Smoking ──→ Cadmium
   ↑
  SES

Age, Sex ──→ Mortality  (covariates, not confounders)
```

The key assumptions encoded in this DAG are:

- **Race** operates as a distal structural cause — it influences mortality not through biology but through the socioeconomic and environmental conditions that racial groups are exposed to. If this assumption is correct, race should lose its independent predictive power once we adjust for the structural pathways (SES and cadmium).
- **SES** is a confounder of the cadmium-mortality relationship: people in more deprived circumstances have both higher cadmium exposure (through residential and occupational pathways) *and* higher mortality (through multiple independent mechanisms including healthcare access and nutrition).
- **Smoking** plays a dual role — it is both a confounder (SES influences smoking, and smoking influences mortality independently) and a mediating pathway (smoking is one route through which people accumulate cadmium).
- **Unmeasured confounders** almost certainly exist: healthcare access, full residential history, psychosocial stress, and dietary quality are not captured in NHANES. This means our causal estimates should be interpreted as upper bounds.

**A note on DAG construction and its relevance to the PhD project.** The DAG in this analysis was specified by the researcher based on existing literature. The proposed Manchester PhD explicitly involves building causal graphs through stakeholder workshops — bringing together clinicians, environmental scientists, and patient representatives to elicit expert knowledge about the causal structure. This participatory approach would strengthen the causal assumptions considerably, and would likely add important nodes not available here (for example, residential proximity to industrial sites, green space access, and specific treatment pathways). The shift from researcher-specified to collaboratively-specified DAGs is one of the key methodological advances the PhD project would enable.

## Hypotheses

| # | Hypothesis |
|---|---|
| H1 | Higher blood cadmium independently associated with all-cause mortality after full adjustment |
| H2 | Lower SES associated with higher blood cadmium (dose-response) |
| H3 | Black Americans have higher cadmium than White Americans within each smoking category |
| H4 | Smoking adjustment attenuates cadmium HR (informal mediation) |
| H5 | Race does not modify the cadmium-mortality association |
| H6 | Equalising cadmium between below-poverty and middle-high SES increases 10-year survival |

---

# Level 1 — Association

*"Can observing patterns in the data tell us something about the relationship between exposure and outcome?"*

At this first level, we are simply describing what we see — are cadmium levels, socioeconomic status, race, and mortality related to each other? No causal claims are made here; the goal is to characterise the data and identify the patterns that motivate the causal questions addressed in Levels 2 and 3.

## Table 1: Baseline Characteristics
```{r table1}
wb_data %>%
  dplyr::select(race, age, sex, ses_cat, smoking,
                cadmium_raw, allcause_death) %>%
  mutate(
    allcause_death = factor(allcause_death,
                            levels = c(0, 1),
                            labels = c("Alive", "Died"))
  ) %>%
  tbl_summary(
    by = race,
    label = list(
      age            ~ "Age (years)",
      sex            ~ "Sex",
      ses_cat        ~ "Socioeconomic status",
      smoking        ~ "Smoking status",
      cadmium_raw    ~ "Blood cadmium (ug/L)",
      allcause_death ~ "All-cause mortality"
    ),
    statistic = list(
      all_continuous()  ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing_text = "Missing"
  ) %>%
  add_overall() %>%
  add_p() %>%
  bold_labels() %>%
  modify_caption(
    "**Table 1. Baseline characteristics by race**
    Pooled NHANES 2007–2012 (White + Black Americans)"
  )
```

**Interpretation:**

- **Age:** White Americans in this sample are older on average (median 52 vs 50 years). This age difference turns out to be critically important — it explains the seemingly paradoxical finding in the crude Kaplan-Meier curves below, where Black Americans appear to have *better* survival. The apparent advantage is entirely an artefact of the younger age distribution.
- **SES:** Black Americans are substantially more socioeconomically deprived, with approximately 25% falling below the poverty line compared to 17% of White Americans. This is the dominant structural difference between the two groups and is central to the environmental justice argument developed throughout this analysis.
- **Smoking:** Despite smoking at lower rates than White Americans (25% vs 23% current smokers), Black Americans carry equal or slightly higher blood cadmium levels (median 0.37 vs 0.35 μg/L). This is the key observation motivating the investigation into non-smoking cadmium sources — if cadmium were simply a proxy for cigarette exposure, we would expect cadmium levels to track smoking prevalence. The fact that they do not points toward industrial, dietary, and residential sources that are distributed along socioeconomic lines.

## Cadmium Distribution
```{r cadmium-plots, fig.height=5}
p1 <- wb_data %>%
  ggplot(aes(race, cadmium_log, fill = race)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  scale_fill_manual(values = c("#2E9FDF", "#E7B800")) +
  labs(title = "Cadmium by Race",
       x = NULL, y = "Log cadmium (ug/L)") +
  theme_minimal() + theme(legend.position = "none")

p2 <- wb_data %>%
  filter(!is.na(ses_cat)) %>%
  ggplot(aes(ses_cat, cadmium_log, fill = ses_cat)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  scale_fill_manual(values = c("#2E9FDF", "#E7B800", "#E84B23")) +
  labs(title = "Cadmium by SES",
       x = NULL, y = "Log cadmium (ug/L)") +
  theme_minimal() + theme(legend.position = "none")

p3 <- wb_data %>%
  filter(!is.na(smoking)) %>%
  ggplot(aes(smoking, cadmium_log, fill = smoking)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  scale_fill_manual(values = c("#2E9FDF", "#E7B800", "#E84B23")) +
  labs(title = "Cadmium by Smoking",
       x = NULL, y = "Log cadmium (ug/L)") +
  theme_minimal() + theme(legend.position = "none")

p1 + p2 + p3
```
```{r cadmium-race-smoking, fig.height=5}
wb_data %>%
  filter(!is.na(smoking)) %>%
  ggplot(aes(race, cadmium_log, fill = race)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  facet_wrap(~smoking) +
  scale_fill_manual(values = c("#2E9FDF", "#E7B800")) +
  labs(
    title    = "Blood Cadmium by Race, Stratified by Smoking",
    subtitle = "Racial cadmium gap persists within each smoking category → non-smoking sources contribute",
    x = NULL, y = "Log cadmium (ug/L)"
  ) +
  theme_minimal() + theme(legend.position = "none")
```

**Plot 2 (Cadmium by SES)** shows the strongest and most important signal — a clear dose-response gradient where each step down in income is associated with progressively higher blood cadmium. This directly supports the hypothesis that socioeconomic deprivation drives environmental toxic burden in a graded manner, not simply as a threshold effect affecting only the most deprived.

**Plot 4 (Race × Smoking stratification)** provides critical evidence for the non-smoking cadmium source argument. Within every smoking category — never smokers, former/never smokers, and current smokers — Black Americans carry equal or higher cadmium levels than White Americans. If smoking were the only meaningful source of cadmium, racial differences in cadmium should disappear when we compare people with the same smoking behaviour. The fact that the gap persists points to industrial emissions, dietary contamination, and residential proximity to pollution sources as independent contributors to cadmium burden.

## Kaplan-Meier Survival Curves
```{r km1, fig.height=7}
km_cad <- survfit(Surv(surv_months, cancer_death) ~ cadmium_hilow,
                  data = pooled_data)
ggsurvplot(
  km_cad, data = pooled_data,
  pval = TRUE, conf.int = TRUE,
  risk.table = TRUE, risk.table.height = 0.25,
  legend.labs  = c("Low cadmium", "High cadmium"),
  legend.title = "",
  title   = "Cancer mortality by cadmium level\nPooled NHANES 2007–2012 (all races)",
  xlab    = "Time (months)", ylab = "Survival probability",
  palette = c("#2166ac", "#d6604d"),
  ggtheme = theme_minimal()
)
```
```{r km2, fig.height=7}
km_race <- survfit(Surv(surv_months, allcause_death) ~ race,
                   data = wb_data)
ggsurvplot(
  km_race, data = wb_data,
  pval = TRUE, conf.int = TRUE,
  risk.table = TRUE, risk.table.height = 0.25,
  legend.labs  = c("White", "Black"), legend.title = "Race",
  title   = "All-cause survival by race — CRUDE (age confounded)\nPooled NHANES 2007–2012",
  xlab    = "Time (months)", ylab = "Survival probability",
  palette = c("#2E9FDF", "#E7B800"),
  ggtheme = theme_minimal()
)
```
```{r km3, fig.height=7}
km_ses <- survfit(
  Surv(surv_months, allcause_death) ~ ses_cat,
  data = wb_data %>% filter(!is.na(ses_cat))
)
ggsurvplot(
  km_ses, data = wb_data %>% filter(!is.na(ses_cat)),
  pval = TRUE, conf.int = FALSE,
  risk.table = TRUE, risk.table.height = 0.25,
  legend.title = "SES",
  title   = "All-cause survival by SES\nPooled NHANES 2007–2012 (White + Black)",
  xlab    = "Time (months)", ylab = "Survival probability",
  palette = c("#2E9FDF", "#E7B800", "#E84B23"),
  ggtheme = theme_minimal()
)
```

**KM Curve 1 (Cancer mortality by cadmium):** The survival curves for high and low cadmium groups begin to separate at around 40 months and diverge progressively thereafter (p<0.0001). This pattern is consistent with cadmium's known cumulative biological mechanism — toxic effects build over time as cadmium accumulates in tissues, so we would expect the survival difference to widen with longer follow-up rather than appearing immediately.

**KM Curve 2 (All-cause survival by race):** This curve illustrates one of the most important methodological lessons in observational epidemiology. In the crude (unadjusted) comparison, Black Americans appear to survive *better* than White Americans. This is entirely misleading — it reflects the fact that the White American sample is older on average. This is a textbook example of why unadjusted comparisons can lead to seriously wrong conclusions, and why the multivariable modelling in Level 2 is essential.

**KM Curve 3 (All-cause survival by SES):** There is a clear socioeconomic gradient in survival, with lower income associated with worse outcomes. One noteworthy pattern is that the below-poverty group appears to survive slightly longer than the low-middle income group — this is the **near-poor paradox**, a well-documented phenomenon in US health data. People just above the poverty threshold earn too much to qualify for Medicaid (the government health insurance programme for low-income Americans) but too little to afford private health insurance, creating a coverage gap that translates into worse health outcomes. This paradox is specific to the US healthcare system and would not be expected in a universal healthcare setting like the UK's NHS — a point directly relevant to the proposed PhD research.

---

# Level 2 — Intervention

*"What would happen if we could change the exposure while holding everything else constant?"*

At this level, we move beyond description to ask what the effect of cadmium on mortality would be if we could intervene — that is, if we could change someone's cadmium level while keeping their other characteristics fixed. In practice, we cannot run randomised experiments assigning people to different cadmium exposures, so we approximate the interventional question using multivariable regression. The key idea is that by sequentially adjusting for confounders identified in the DAG, we can isolate the portion of the cadmium-mortality association that is not explained by other factors.

## Cox Model Sequence

The analytical strategy here is to build a sequence of five Cox proportional hazards models, each adding a new block of covariates. By tracking how the cadmium hazard ratio (HR) changes as we add variables, we can see the confounding structure in action — each change reveals which variables were inflating or masking the true cadmium effect.
```{r cox-models}
m1 <- coxph(Surv(surv_months, allcause_death) ~
              cadmium_log,
            data = wb_data)

m2 <- coxph(Surv(surv_months, allcause_death) ~
              cadmium_log + age + sex,
            data = wb_data)

m3 <- coxph(Surv(surv_months, allcause_death) ~
              cadmium_log + age + sex + race + ses_cat,
            data = wb_data)

m4 <- coxph(Surv(surv_months, allcause_death) ~
              cadmium_log + age + sex + race + ses_cat + smoking,
            data = wb_data)

m5 <- coxph(Surv(surv_months, allcause_death) ~
              cadmium_log + age + sex + race + ses_cat + smoking +
              cadmium_log:race,
            data = wb_data)
```

## HR Attenuation Pattern
```{r hr-attenuation}
models     <- list(m1, m2, m3, m4, m5)
mod_labels <- c("Model 1: Crude",
                "Model 2: + Age + Sex",
                "Model 3: + Race + SES",
                "Model 4: + Smoking (FINAL)",
                "Model 5: + Cadmium × Race")

cat("Cadmium HR across sequential models\n")
cat("(per e-fold increase in blood cadmium)\n")
cat("=========================================\n\n")

hrs <- c()
for (i in seq_along(models)) {
  r <- tidy(models[[i]], exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term == "cadmium_log")
  hrs[i] <- r$estimate
  cat(mod_labels[i], "\n")
  cat("  HR =", round(r$estimate, 3),
      "| 95% CI:", round(r$conf.low, 3), "–", round(r$conf.high, 3),
      "| p =", ifelse(r$p.value < 0.001, "<0.001",
                      round(r$p.value, 3)), "\n\n")
}

cat("=========================================\n")
cat("Attenuation decomposition:\n\n")
cat("M1→M2 (+age/sex):  ",
    round((hrs[2]-hrs[1])/hrs[1]*100, 1),
    "% — negative confounding by age\n")
cat("M2→M3 (+race/SES): ",
    round((hrs[3]-hrs[2])/hrs[2]*100, 1),
    "% — positive confounding by SES\n")
cat("M3→M4 (+smoking):  ",
    round((hrs[4]-hrs[3])/hrs[3]*100, 1),
    "% — smoking pathway attenuated\n")
cat("Final independent HR:", round(hrs[4], 3), "\n")
```

**Why not a formal mediation analysis?** It is tempting to interpret the HR changes across models as evidence of mediation — for example, concluding that smoking "mediates" a certain percentage of the cadmium effect because the HR drops when smoking is added. However, the Cox hazard ratio has a mathematical property called **non-collapsibility**: HRs can change when covariates are added even when there is no true mediation, purely as a mathematical artefact of the model structure (VanderWeele 2011; Lange & Hansen 2011). This means that the naive "proportion mediated" calculation would mix together true mediation with statistical noise. We therefore present the HR attenuation pattern here as *indicative structural evidence* — it shows us the direction and approximate magnitude of confounding — but reserve the formal causal estimate for the g-computation analysis in Level 3, which operates on the survival probability scale where this non-collapsibility problem does not arise.

## Table 2: Full Model 4 Results
```{r model4-table}
tidy(m4, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(
    p_fmt = case_when(
      p.value < 0.001 ~ "<0.001 ***",
      p.value < 0.01  ~ paste0(round(p.value, 3), " **"),
      p.value < 0.05  ~ paste0(round(p.value, 3), " *"),
      TRUE            ~ as.character(round(p.value, 3))
    )
  ) %>%
  dplyr::select(term, estimate, conf.low, conf.high, p_fmt) %>%
  mutate(across(c(estimate, conf.low, conf.high), ~round(., 3))) %>%
  kable(
    col.names = c("Variable", "HR", "Lower 95% CI",
                  "Upper 95% CI", "p-value"),
    caption = "**Table 2. Final adjusted Cox model (Model 4)**
    All-cause mortality, White + Black Americans, pooled NHANES 2007–2012"
  )
```

**Key findings from the final model:**

**Cadmium (HR 1.364, p<0.001):** After accounting for age, sex, race, SES, and smoking, each e-fold (approximately 2.72-fold) increase in blood cadmium is associated with 36% higher all-cause mortality. This is a substantial and highly significant independent effect, confirming that cadmium is not merely a proxy for smoking or poverty — it contributes to mortality through its own biological pathways (nephrotoxicity, carcinogenesis, cardiovascular damage). **H1 confirmed.**

**Race (Black vs White, HR 0.938, p = 0.282):** This is perhaps the most important finding in the entire analysis. After adjusting for the structural pathways — SES, cadmium exposure, and smoking — race is no longer a significant predictor of mortality. The crude racial gap in survival that appeared in the Kaplan-Meier curves is entirely explained by measurable structural factors. This is consistent with the interpretation that racial health inequalities in this population are produced by the conditions in which people live — residential segregation leading to SES disadvantage, leading to higher environmental toxic burden — rather than by inherent biological differences between racial groups.

**SES — the dominant driver:** Below-poverty status carries a hazard ratio of approximately 2.40, meaning 140% higher mortality compared to the middle-high income reference group. This effect dwarfs the cadmium effect, which is expected — cadmium is one specific downstream pathway through which deprivation kills, not the only one. The fact that SES remains so strongly predictive even after cadmium adjustment tells us that many other deprivation-related mechanisms (healthcare access, nutrition, housing quality, psychosocial stress) are also operating simultaneously.

## Interaction Test
```{r interaction}
tidy(m5, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(grepl("cadmium", term)) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  kable(caption = "**Cadmium × Race interaction (Model 5)**")
```

The interaction term is not statistically significant (p ≈ 0.61), meaning the relationship between cadmium and mortality does not differ meaningfully between Black and White Americans. Cadmium appears to be equally harmful regardless of race — the disparity lies in who gets exposed, not in how the exposure affects health. **H5 confirmed.** This finding has important implications: it means that interventions reducing cadmium exposure would benefit all racial groups equally, but would disproportionately help Black Americans because they currently carry higher exposure levels.

## Schoenfeld Diagnostics
```{r schoenfeld, fig.height=8}
ph_test <- cox.zph(m4)
print(ph_test)
par(mfrow = c(3, 3))
plot(ph_test)
par(mfrow = c(1, 1))
```

The Cox proportional hazards model assumes that the ratio of hazards between groups remains constant over time. The Schoenfeld residual test checks this assumption for each variable. Cadmium passes cleanly (p = 0.665), meaning its effect on mortality is consistent across the follow-up period. Age is flagged as borderline (p = 0.017), but visual inspection of the Schoenfeld residual plots shows no systematic trend — this is likely an overpowered test, as with N ≈ 10,000 even trivially small departures from proportionality become statistically detectable. The global test is also non-significant (p = 0.053). The proportional hazards assumption is satisfied for all practical purposes.

## Cotinine Sensitivity Analysis

A natural concern with using cadmium as a biomarker of environmental toxic burden is that it might simply be measuring smoking exposure. To test this, we compare cadmium against **serum cotinine** — a metabolite of nicotine that is exclusively a marker of tobacco smoke exposure. If cadmium's mortality association were entirely driven by smoking, we would expect cotinine to show a similar or stronger effect. The comparison below shows that cadmium (HR 1.364) has a substantially larger effect on mortality than cotinine after full adjustment, confirming that non-smoking environmental sources of cadmium — industrial emissions, dietary contamination, occupational exposure — contribute independently to the mortality risk.
```{r cotinine}
m_cot_crude <- coxph(Surv(surv_months, allcause_death) ~
                       cotinine_log, data = wb_data)
m_cot_adj   <- coxph(Surv(surv_months, allcause_death) ~
                       cotinine_log + age + sex +
                       race + ses_cat + smoking, data = wb_data)

cat("Cotinine crude HR:\n")
tidy(m_cot_crude, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term == "cotinine_log") %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  dplyr::select(term, estimate, conf.low, conf.high, p.value) %>%
  print()

cat("\nCotinine adjusted HR:\n")
tidy(m_cot_adj, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term == "cotinine_log") %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  dplyr::select(term, estimate, conf.low, conf.high, p.value) %>%
  print()

cat("\nCadmium adjusted HR (Model 4):",
    round(exp(coef(m4)["cadmium_log"]), 3), "\n")
cat("\nCotinine captures smoking pathway only.\n")
cat("Cadmium captures smoking + industrial + dietary + residential burden.\n")
cat("Larger cadmium HR confirms non-smoking sources contribute independently.\n")
```

---

# Level 3 — Counterfactual

*"What would have happened if things had been different?"*

This is the most ambitious level of causal reasoning. Rather than asking whether cadmium and mortality are associated (Level 1) or what happens when we statistically adjust for confounders (Level 2), we now ask a directly policy-relevant question: **if we could change the cadmium exposure of deprived communities to match that of affluent communities, how many lives would be saved?**

This is a counterfactual question — it asks about a world that does not exist but could be brought into existence through environmental regulation, industrial pollution controls, or housing policy.

## Causal Question

> *"What would the 10-year survival of people below the poverty line be if their blood cadmium levels matched those of middle-high income people?"*

To answer this, we use **g-computation** — a method from the causal inference literature that allows us to simulate what would happen under a hypothetical intervention. The approach works as follows: first, we fit a Weibull accelerated failure time (AFT) model that predicts survival as a function of cadmium, age, sex, race, and smoking. Then, we take the below-poverty group, replace their cadmium values with values randomly drawn from the middle-high income group, and predict their survival under this counterfactual exposure distribution.

An important modelling decision: `ses_cat` is deliberately *excluded* from the outcome model. This is because we want to isolate the specific pathway from SES through cadmium to mortality. If we included SES in the outcome model, it would absorb the very effect we are trying to quantify — the portion of SES-related mortality that operates through the cadmium pathway. In other words, we are asking: "of all the ways that poverty kills, how much goes through the cadmium channel specifically?"

## G-Computation
```{r gcomp}
poor_dat <- wb_data %>%
  filter(ses_cat == "Below poverty",
         !is.na(cadmium_log), !is.na(age),
         !is.na(sex), !is.na(smoking))

rich_dat <- wb_data %>%
  filter(ses_cat == "Middle-high",
         !is.na(cadmium_log), !is.na(age),
         !is.na(sex), !is.na(smoking))

cat("Group sizes:\n")
cat("  Below poverty: n =", nrow(poor_dat), "\n")
cat("  Middle-high:   n =", nrow(rich_dat), "\n")
cat("\nCadmium gap (log scale):\n")
cat("  Below poverty median:", round(median(poor_dat$cadmium_log), 3), "\n")
cat("  Middle-high median:  ", round(median(rich_dat$cadmium_log), 3), "\n")

gcomp_mod <- survreg(
  Surv(surv_months, allcause_death) ~
    cadmium_log + age + sex + race + smoking,
  data = wb_data, dist = "weibull"
)

pred_surv <- function(mod, dat, t) {
  lp <- predict(mod, newdata = dat, type = "lp")
  1 - pweibull(t, shape = 1 / mod$scale, scale = exp(lp))
}

surv_obs <- pred_surv(gcomp_mod, poor_dat, 120)

set.seed(42)
poor_cf <- poor_dat %>%
  mutate(cadmium_log = sample(rich_dat$cadmium_log,
                               size = n(), replace = TRUE))
surv_cf <- pred_surv(gcomp_mod, poor_cf, 120)

obs_10yr  <- mean(surv_obs, na.rm = TRUE)
cf_10yr   <- mean(surv_cf,  na.rm = TRUE)
abs_diff  <- cf_10yr - obs_10yr
attr_frac <- abs_diff / (1 - obs_10yr) * 100

cat("\nPoint estimates (10-year survival):\n")
cat("  Observed (below poverty): ", round(obs_10yr * 100, 2), "%\n")
cat("  Counterfactual:           ", round(cf_10yr  * 100, 2), "%\n")
cat("  Absolute difference:      ", round(abs_diff * 100, 2), "pp\n")
cat("  Attributable fraction:    ", round(attr_frac, 2), "%\n")
```

## Bootstrap Confidence Intervals
```{r bootstrap, cache=TRUE}
set.seed(42)
n_boot     <- 500
boot_diffs <- numeric(n_boot)

for (b in seq_len(n_boot)) {
  bd <- wb_data[sample(nrow(wb_data), replace = TRUE), ]
  bp <- bd %>% filter(ses_cat == "Below poverty",
                      !is.na(cadmium_log), !is.na(age),
                      !is.na(sex), !is.na(smoking))
  br <- bd %>% filter(ses_cat == "Middle-high",
                      !is.na(cadmium_log), !is.na(age),
                      !is.na(sex), !is.na(smoking))
  if (nrow(bp) < 50 || nrow(br) < 50) next
  bm <- tryCatch(
    survreg(Surv(surv_months, allcause_death) ~
              cadmium_log + age + sex + race + smoking,
            data = bd, dist = "weibull"),
    error = function(e) NULL
  )
  if (is.null(bm)) next
  s_obs <- mean(pred_surv(bm, bp, 120), na.rm = TRUE)
  bp_cf <- bp %>% mutate(
    cadmium_log = sample(br$cadmium_log, size = n(), replace = TRUE)
  )
  s_cf <- mean(pred_surv(bm, bp_cf, 120), na.rm = TRUE)
  boot_diffs[b] <- s_cf - s_obs
}

boot_diffs <- boot_diffs[boot_diffs != 0]
ci_lo <- quantile(boot_diffs, 0.025) * 100
ci_hi <- quantile(boot_diffs, 0.975) * 100

cat("=================================================\n")
cat("G-COMPUTATION FINAL RESULT\n")
cat("=================================================\n")
cat("Observed 10-year survival (below poverty):",
    round(obs_10yr * 100, 2), "%\n")
cat("Counterfactual 10-year survival:          ",
    round(cf_10yr  * 100, 2), "%\n")
cat("Absolute difference:",
    round(abs_diff * 100, 2), "pp",
    "(95% CI:", round(ci_lo, 2), "–", round(ci_hi, 2), "pp)\n")
cat("Attributable fraction:", round(attr_frac, 2), "%\n")
cat("Bootstrap iterations: ", length(boot_diffs), "\n")
cat("=================================================\n")
```

Equalising cadmium exposure would increase 10-year survival among the below-poverty group by approximately **1.62 percentage points (95% CI: 1.14–2.14)**. To put this in context, the attributable fraction of approximately 12.85% means that roughly **one in eight deaths** among people below the poverty line can be attributed to their disproportionately high cadmium exposure compared to middle-high income individuals. The 95% confidence interval excludes zero, indicating that this finding is robust across 500 bootstrap replications.

While 1.62 percentage points may sound modest in absolute terms, it is important to remember that this represents the contribution of a *single* environmental pollutant through a *single* biomarker. The total environmental toxic burden includes hundreds of other pollutants (PM2.5, nitrogen oxides, lead, volatile organic compounds) that are also distributed along socioeconomic lines. The cumulative effect of addressing multiple environmental exposures simultaneously would likely be substantially larger — a hypothesis that the proposed Manchester PhD is designed to test using multi-pollutant exposure data from the Natural Environment Research Council (NERC).

**Required causal assumptions:** For the g-computation estimate to have a causal interpretation, three assumptions must hold: (1) **Consistency** — the observed outcomes under each exposure level correspond to what would happen under an intervention setting that level; (2) **Exchangeability** — there are no unmeasured confounders that differ between SES groups beyond those we have adjusted for; and (3) **No interference** — one person's cadmium level does not affect another person's survival. Assumption (2) is almost certainly partially violated (healthcare access, residential history, and psychosocial stress are unmeasured), which means our attributable fraction should be interpreted as an **upper bound** on the true causal effect.

---

# Results Summary
```{r results-table}
tibble(
  Level = c(
    rep("Level 1 — Association", 3),
    rep("Level 2 — Intervention", 5),
    "Level 3 — Counterfactual"
  ),
  Analysis = c(
    "KM: Cancer mortality by cadmium",
    "KM: All-cause survival by race (crude)",
    "KM: All-cause survival by SES",
    "Cox Model 1 — crude cadmium",
    "Cox Model 2 — + age, sex",
    "Cox Model 3 — + race, SES",
    "Cox Model 4 — + smoking (final)",
    "Race × Cadmium interaction (Model 5)",
    "G-computation: SES cadmium equalisation"
  ),
  Finding = c(
    "High cadmium → worse cancer survival (p<0.0001)",
    "Black > White crude — age confounding",
    "Clear SES gradient; near-poor paradox",
    "HR = 1.439 (1.365–1.516), p<0.001",
    "HR = 1.583 — age negative confounding",
    "HR = 1.493 — SES positive confounding",
    "HR = 1.364 — race NS (p≈0.28); SES HR ~2.40",
    "HR ≈ 0.96, p≈0.61 — no effect modification",
    "1.62pp gain (95% CI 1.14–2.14); AF ~12.85%"
  )
) %>%
  kable(caption = "**Table 3. Summary of findings**")
```

---

# Discussion

## Principal Findings

This analysis of 10,410 White and Black Americans from three pooled NHANES cycles demonstrates three principal findings.

**First**, blood cadmium is independently associated with 36% higher all-cause mortality per e-fold increase (HR 1.364, 95% CI 1.247–1.492) after adjusting for age, sex, race, SES, and smoking. The cotinine sensitivity analysis provides additional confidence in this finding — because cotinine captures only the smoking-related portion of toxic exposure while cadmium captures both smoking and non-smoking sources, the substantially larger cadmium effect confirms that industrial, dietary, and residential environmental exposures contribute to mortality independently of tobacco use.

**Second**, the sequential model-building approach reveals the confounding structure with considerable clarity. Age acts as a *negative* confounder — younger people tend to have higher cadmium (likely through occupational exposure) but lower mortality, so failing to adjust for age actually masks part of the cadmium effect. SES acts as a *positive* confounder — deprived people have both higher cadmium and higher mortality through independent pathways, inflating the crude association. Smoking adjustment removes the portion of cadmium's effect that operates specifically through the cigarette inhalation pathway, leaving the residual effect attributable to non-smoking environmental sources. Understanding this confounding architecture is not just a statistical exercise — it tells us which policy levers would be effective. If cadmium's effect were entirely through smoking, the policy response would be tobacco control; the fact that a substantial portion operates through non-smoking pathways points instead toward environmental regulation, industrial emission standards, and land-use planning.

**Third**, race is not a significant predictor of mortality after adjustment for structural pathways (HR 0.938, p = 0.282). This finding carries important implications. The crude racial mortality gap — which might be interpreted as evidence of biological differences between racial groups — is entirely accounted for by measurable structural factors: age composition, socioeconomic status, cadmium exposure, and smoking patterns. This is consistent with a growing body of evidence that racial health inequalities are produced by the unequal social and environmental conditions in which different racial groups live, rather than by inherent biological vulnerability.

## Environmental Justice Implications

The g-computation finding translates the statistical results into a directly policy-relevant quantity: approximately 12.85% of excess mortality among people below the poverty line is attributable to their disproportionately high cadmium exposure. This is a conservative estimate given the single-pollutant, single-biomarker design of this analysis.

An important nuance emerges from the SES-cadmium dose-response pattern: environmental toxic burden increases in a graded fashion with each step down in income, not as a threshold effect affecting only the most deprived. This means that policies targeting only the poorest communities — while necessary — would miss the substantial excess burden carried by the low-middle income group. Effective environmental health policy needs to address the entire socioeconomic gradient.

## Limitations

This analysis has several important limitations, each of which points toward specific methodological improvements that the proposed Manchester PhD is designed to address:

**Single timepoint cadmium measurement.** NHANES measures cadmium at a single survey visit. Although cadmium's long biological half-life means a single measurement is a reasonable proxy for cumulative exposure, it still introduces measurement error that likely attenuates our estimates toward the null — meaning the true effect is probably larger than what we observe.

**All-cause mortality as the outcome.** Using all-cause mortality dilutes the cadmium-specific signal because many deaths are caused by conditions unrelated to cadmium exposure. Cancer-specific analysis would be more informative, but the current sample provides insufficient cancer deaths (425 total, far fewer within subgroups) for stable site-specific modelling.

**Unmeasured confounding.** Healthcare access, full residential history, dietary quality, and psychosocial stress are not available in NHANES but likely differ between SES groups in ways that independently affect mortality. The attributable fraction should therefore be interpreted as an upper bound.

**Cross-sectional exposure assessment.** Because cadmium is measured at the same timepoint as baseline characteristics, we cannot fully exclude the possibility that the causal direction is partially reversed — though cadmium's decades-long accumulation makes reverse causality less plausible than it would be for a rapidly metabolised biomarker.

**US-specific context.** The NHANES dataset reflects the US healthcare and social system. The Medicaid eligibility cliff that produces the near-poor paradox in the SES-survival curves would not be expected in the UK's universal healthcare system. Generalising these findings to the UK requires caution — and direct UK-based analysis.

**Restriction to White and Black Americans.** This analysis focuses on White and Black Americans due to statistical power constraints. The ethnic groups most relevant to health inequality research in the UK (South Asian, Black Caribbean, Black African communities) are different from those studied here, and the environmental exposure patterns, residential geography, and healthcare access pathways will differ accordingly.

## Bridge to the Proposed PhD Research

The limitations outlined above are not merely caveats — they represent precisely the methodological gaps that the proposed Manchester PhD project is designed to address. The table below maps each limitation of the current analysis to the specific capability that the PhD project offers:

| Limitation in this analysis | How the PhD project addresses it |
|---|---|
| Single cadmium biomarker at one timepoint | Postcode-level NERC atmospheric deposition data provides spatially resolved, longitudinal measurements of multiple pollutants (PM2.5, PM10, NOx, cadmium, and others) linked to residential history |
| All-cause mortality dilutes signal | The Christie NHS Foundation Trust cancer registry enables cancer site-specific outcome analysis — distinguishing lung, kidney, bladder, and other cancers with different environmental exposure profiles |
| Underpowered cancer-specific analysis | Linking a large specialist cancer centre's complete patient records provides sufficient events for site-specific and treatment-response modelling |
| Unmeasured confounders | Availability of treatment data (chemotherapy regimens, radiotherapy, surgical outcomes) and green space / noise pollution data enables adjustment for pathways not measurable in NHANES |
| US healthcare system specifics | Situating the analysis within the NHS eliminates the Medicaid eligibility cliff confounding and provides a universal-access baseline, allowing cleaner isolation of environmental effects |
| Researcher-specified DAG | The PhD's stakeholder workshop approach — bringing together clinicians, environmental scientists, and patient representatives — enables participatory construction of causal graphs grounded in expert knowledge from multiple disciplines |
| Single-pollutant analysis | NERC multi-pollutant data enables examination of cumulative and interactive effects of multiple environmental exposures, better reflecting real-world conditions where communities face mixtures of pollutants rather than isolated chemicals |

Beyond addressing these specific limitations, the PhD offers three broader methodological advances. First, the shift from a blood biomarker (which captures individual-level exposure from all sources combined) to geographically referenced pollution data (which captures place-based exposure) enables investigation of environmental justice mechanisms at the neighbourhood level — asking not just whether deprived people have higher exposure, but *where* and *why*. Second, the inclusion of treatment and outcome data (survival, treatment toxicity, quality of life) enables analysis of whether environmental pollutants affect not only cancer incidence but also how well patients respond to treatment once diagnosed — a question that current environmental health research has largely not addressed. Third, the causal inference framework demonstrated here — DAG specification, sequential confounding decomposition, g-computation for counterfactual estimation — can be extended in the PhD to incorporate more advanced methods such as targeted maximum likelihood estimation (TMLE) for doubly-robust inference, or instrumental variable approaches using geographic variation in pollution as natural experiments.

This analysis demonstrates that I can work with complex linked datasets, implement a structured causal inference framework, and communicate findings in terms of their policy implications — skills that are directly transferable to the proposed PhD research programme.

---

# Conclusion

Blood cadmium is independently associated with 36% higher all-cause mortality (HR 1.364, 95% CI 1.247–1.492) after full adjustment for demographic, socioeconomic, and smoking confounders. The racial mortality gap disappears entirely after structural pathway adjustment, consistent with the interpretation that racial health inequalities in this population are driven by socioeconomic and environmental mechanisms rather than biological differences. Socioeconomic deprivation is the dominant mortality driver (HR ~2.40 for below-poverty status), with cadmium representing one specific environmental pathway through which deprivation translates into excess death.

The g-computation counterfactual analysis estimates that equalising cadmium exposure between the most and least deprived groups would improve 10-year survival by 1.62 percentage points (95% CI 1.14–2.14), with approximately 12.85% of excess mortality in deprived communities attributable to differential cadmium burden. As an upper bound estimate from a single pollutant, this finding underscores the potential scale of health gains achievable through environmental regulation and industrial pollution policy — and motivates the multi-pollutant, cancer-specific, UK-based analysis proposed in the Manchester PhD programme.

# Generative AI Use

This analysis was developed with the assistance of Claude (Anthropic), 
a large language model, used as an interactive analytical and writing 
collaborator. Specifically, Claude was used to discuss causal inference 
methodology (Pearl's ladder of causation, g-computation, HR 
non-collapsibility), draft and iteratively refine R code for data 
harmonisation, Cox regression, and bootstrap confidence intervals, and 
assist in structuring the narrative interpretation of findings.

All statistical outputs, model results, and numerical findings reported 
in this document were verified by the author against R console output. 
The causal reasoning, epidemiological interpretation, and conclusions 
reflect the author's own judgement and domain knowledge. All code was 
reviewed, tested, and executed by the author in a local R environment 
before inclusion in this report.

The use of AI assistance is declared in the interest of transparency 
and consistent with emerging norms for responsible AI use in academic 
and research contexts.

---

# Session Information
```{r session-info}
sessionInfo()
```